---
title: "Basic Model"
date: "07/12/2015"
output: pdf_document
---

## 1. Notations

- $n_s$ is the number of sites
- $N_{ts}$ be the number of species at time $t$ and site $s$
- $d_{si}$ be the distance between two sites $s$ and site $i$. Currently I am using the signed(North is positive) distance(unit is degree) of the latitude between each two sites.
- The environmental variables are $e_{chi}, e_{sst}$ and $e_{upw}$
- Let $y_{ts}=\log N_{(t+1)s}$.

## 2. The model

The dispersal kernel(for site $i$) is some constant($K=e^{k}$) times the following term:
\[
N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}},
\]

where $N_{ti}$ is the number of species at time $t$ and site $i$.

So, the raw model is:

\[
y=k+\log(\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}})+\beta_1e_{chi}+\beta_2e_{sst}+
\beta_3e_{upw}+\epsilon,
\]


where $\epsilon\sim N(0,\sigma^2)$ .  So we have 

- three parameters related to the $\textbf{dispersal}$: $k, \mu_d, \sigma_d$.
- three parameters for the $\textbf{environmental}$ variables $\beta_1, \beta_2, \beta_3$.

## 3. Estimation

Let $\theta=(k, \mu_d, \sigma_d, \beta_1, \beta_2, \beta_3)$, $x$ be our data ($N$:number of species, $d$:distance between the sites, $e$: the environmental variables), $y=f(x,\theta)$ is our model. Then $y\sim N(f(x,\theta),\sigma^2)$


\[
L(\theta;y,x)=\prod_{j=1}^m \phi(\frac{y_j-f(x_j,\theta)}{\sigma})
\]


and


\[
l(\theta;y,x)=\log L(\theta;y,x)=\sum_{j=1}^m \log \phi(\frac{y_j-f(x_j,\theta)}{\sigma})
\]


**Note**: The model is a nonlinear regression problem. After some calculation about the above log-likelihood function, we can show that: maximum likelihood estimation is equivalent to the nonlinear least square estimation. So we need to minimize the objective function:


\[
\sum_{j=1}^n (y_j-f(x_j,\theta))^2
\]


## 4. About the Data

As for our data, we have 4 years(year0, year1, year2, year3) data. So $y$ should be


\[
y=N_{ts}\quad for\quad [year1(s1,...,s48),year2(s1,...,s48),year3(s1,...,s48)]
\]


Predictive variables $x$ contains three parts: 

1. species number $N$, 

\[
N=N_{ts}\quad for\quad [year0(s1,...,s48),year1(s1,...,s48),year2(s1,...,s48)]
\]

2. the distance $d$, which should be a distance matrix between the sites (s1,...,s48)

3. the environmental variables $e_i$, which is

\[
e_i=e_{i}\quad for\quad [year0(s1,...,s48),year1(s1,...,s48),year2(s1,...,s48)]
\]

## 5. Code and Results

### 5.1 EDA

```{r}
library(synchrony)
data(pisco.data)
head(pisco.data)
summary(pisco.data)
```

In the above data:


- lat and lon: the location of the stations(48 different stations)
- chl, sst, and upwelling: the environmental variables
- mussel_abund: the variable we want to predict
- year: 2000-2004

Now we extract the target variable, and the predictive variables:

```{r}
y=subset(pisco.data,year>2000,select=c(mussel_abund)) # species number
y[y[,1]==0,]=min(y[y[,1]!=0,])/2 # replace 0 to half of the minimum positive number
y=log(y) # take log of the species number

N=subset(pisco.data,year<2003,select=c(mussel_abund)) # (past) species number
#D=coord2dist(pisco.data[1:48,1:2],lower.tri = F) # site distance using lat and lon
D=dist(pisco.data[1:48,1],diag = T,upper = T) # site distance using only (signed) lat
D=as.matrix(D)
for(i in 1:48){
    for(j in 1:48){
        if(i>j){
            D[i,j]=-D[i,j]
        }
    }
}
E=subset(pisco.data,year<2003,select=c(chl,sst,upwelling)) # Environment variable
D=as.matrix(D)
D[1:6,1:6]
```

Here, we define the objective function. Since we want to do minimization(not max), the objective function is the either the negative log-likelihood function, or the nonlinear least squre error(they are equivalent in our case). The input arguments are:

- t: all the parameters which need to do minimization, $(k, \mu_d, \sigma_d, \beta_1, \beta_2, \beta_3)$
- y: target variable(log of species number)
- N: past species number
- D: distance matrix among 48 sites
- E: 3 environmental variables


### 5.2 First Model

\[
y=k+\log(\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}})+\beta_1e_{chi}+\beta_2e_{sst}+
\beta_3e_{upw}+\epsilon,
\]


```{r cache=TRUE}
# log-likelihood function
logl2=function(t,y,N,D,E){
  n=dim(y)[1]
  f=rep(0,n) # value of the regression function
  yr=rep(0:2,each=48) # 3 years(0-2) used: 00-03 for X, 01-04 for y
  for(j in 1:n){
    # dispersal kernal (for the 48 sites)
    Ker=exp(-(D[j%%48+((j%%48)==0)*48,]-t[2])^2/(t[3]^2)) 
    # number of species for 48 sites
    Nj=N[(yr[j]*48+1):((yr[j]+1)*48),] 
    # dispersal + three environment terms
    f[j]=t[1]+log(sum(Ker*Nj))+t[4]*E[j,1]+t[5]*E[j,2]+t[6]*E[j,3] 
  }
  
  return(sum((y[,1]-f)^2)) # return the objective function
}


# do the optimization to find the parameters
t0=c(1,0,1,0.1,0.1,0.1)
res2=nlm(logl2,t0,hessian=T,print.level=1,y=y,N=N,D=D,E=E,iterlim=1e4,steptol=1e-5)
```

```{r}
t=res2$estimate
print("The estimated parameters are:")
print(t)
n=dim(y)[1]
  f=rep(0,n) # value of the regression function
  yr=rep(0:2,each=48) # 3 years(0-2) used: 00-03 for X, 01-04 for y
  for(j in 1:n){
    # dispersal kernal (for the 48 sites)
    Ker=exp(-(D[j%%48+((j%%48)==0)*48,]-t[2])^2/(t[3]^2)) 
    # number of species for 48 sites
    Nj=N[(yr[j]*48+1):((yr[j]+1)*48),] 
    # dispersal + three environment terms
    f[j]=t[1]+log(sum(Ker*Nj))+t[4]*E[j,1]+t[5]*E[j,2]+t[6]*E[j,3] 
  }
MSE=mean((y[,1]-f)^2)
print(paste("MSE in log scale is: ",MSE))
par(mfrow=c(1,1))
options(repr.plot.width = 10)
options(repr.plot.height = 5)
plot(f,y[,1],main=expression(paste("log scale, ", hat(y)," vs y")))
abline(a=0,b=1,col=2)
plot(f,y[,1]-f,main=expression(paste("log scale, ", hat(y)," vs e")))
abline(a=0,b=0,col=2)
```


### 5.3 95\% Confidence Interval for $\mu_d$

Using the fact that MLE is aymptotic normal, $$\sqrt{n}(\hat{\theta}-\theta)\to N(0,\frac{1}{I_1(\theta)}),$$

i.e. $\hat{\theta}=N(\theta,\frac{1}{I_n(\theta)})$, where $I_n(\theta)=nI_1(\theta)$ is the Fisher Information for n sample points, and one sample point. So the 95\% C.I. for $\theta$ is $$\hat{\theta}\pm1.96\frac{1}{\sqrt{J_n(\hat{\theta})}},$$

where $J_n(\hat{\theta})$ is the observed Fisher Information $J_n(\hat{\theta})=-l_n''(\theta)=-\sum_{i=1}^n(\log f(X_i;\theta))''$

$$
\begin{aligned}
l(\theta;y,X)&= \sum_{j=1}^m\log \phi(\frac{y_j-f(x_j;\theta)}{\sigma})\\
 &= \sum_{j=1}^m[\log(\frac{1}{\sqrt{2\pi}})-\frac{1}{2}[\frac{y_j-f(x_j;\theta)}{\sigma}]^2\\
 &=m\log(\frac{1}{\sqrt{2\pi}})-\frac{1}{2\sigma^2}\sum_{j=1}^m[y_j-f(x_j;\theta)]^2
\end{aligned}
$$

$$
\frac{\partial l}{\partial\mu_d}=\frac{1}{\sigma^2}\sum_{j=1}^m[y_j-f(x_j;\theta)]\frac{\partial f}{\partial\mu_d}
$$

$$
\begin{aligned}
\frac{\partial^2 l}{\partial\mu_d^2}&=\frac{1}{\sigma^2}\sum_{j=1}^m\{[-\frac{\partial f}{\partial\mu_d}]\frac{\partial f}{\partial\mu_d}+[y-f]\frac{\partial^2 f}{\partial\mu_d^2}\}\\
&=\frac{1}{\sigma^2}\sum_{j=1}^m[-\frac{\partial f}{\partial\mu_d}-(\frac{\partial f}{\partial\mu_d})^2+[y-f]\frac{\partial^2 f}{\partial\mu_d^2}]
\end{aligned}
$$

where,

$$
\frac{\partial f}{\partial\mu_d}=\frac{1}{\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}}\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}\frac{2(d_{si}-\mu_d)}{\sigma_d^2}
$$

$$
\begin{aligned}
\frac{\partial^2 f}{\partial\mu_d^2}&=\frac{[\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}\frac{2(d_{si}-\mu_d)}{\sigma_d^2}]'[\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}]-[\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}]'[\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}\frac{2(d_{si}-\mu_d)}{\sigma_d^2}]}{[\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}]^2}
\end{aligned}
$$

Here,

$$
[\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}]'=\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}\frac{2(d_{si}-\mu_d)}{\sigma_d^2}
$$


$$
[\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}\frac{2(d_{si}-\mu_d)}{\sigma_d^2}]'=
\sum_{i=1}^{n_s}N_{ti}[e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}(\frac{2(d_{si}-\mu_d)}{\sigma_d^2})^2+e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}}(-\frac{2}{\sigma_d^2})]
$$




```{r}
mu_mle=t[2]
sigma=sd(y[,1]-f)
n=dim(y)[1]
f=rep(0,n) # value of the regression function
fp=rep(0,n)
fpp=rep(0,n)
yr=rep(0:2,each=48) # 3 years(0-2) used: 00-03 for X, 01-04 for y
for(j in 1:n){
    # dispersal kernal (for the 48 sites)
    Ker=exp(-(D[j%%48+((j%%48)==0)*48,]-t[2])^2/(t[3]^2)) 
    # number of species for 48 sites
    Nj=N[(yr[j]*48+1):((yr[j]+1)*48),] 
    # dispersal + three environment terms
    f[j]=t[1]+log(sum(Ker*Nj))+t[4]*E[j,1]+t[5]*E[j,2]+t[6]*E[j,3] 
    #-------------------------------------------------------------
    dmm=D[j%%48+((j%%48)==0)*48,]-t[2]# d_si - mu_d   
    fp[j]=sum(Nj*Ker*2*dmm/t[3]^2)/sum(Ker*Nj) # f'    
    Kp=sum(Nj*(Ker*4*dmm^2-Ker*2/t[3]^2)) # derivative of top for f'    
    kp=sum(Nj*Ker*2*dmm/t[3]^2) # derivative of bottom for f'    
    fpp[j]=(Kp*sum(Ker*Nj)-kp*sum(Nj*Ker*2*dmm/t[3]^2))/sum(Ker*Nj)^2 # f''  
}
lpp=sum(-fp-fp^2+(y-f)*fp)/sigma^2 #l''(wrt mu_d)
mu_Jn=-lpp # observed Fisher Information
mu_l=1.96/mu_Jn # 95% CI half length
c(t[2]-mu_l, t[2]+mu_l) # 95% CI
```


## 6. Comments

- Using $log(N)$ to avoid negative fitting results
- Using signed lat distance to make $\mu_d$ having the meaning that the center of the dispersal kernel(moving towards the North ot South of the site). As we can see from the fitting result(second parameter), $\mu_d$ is about 0.0035 degree to the South.
- Zero values(in this data set, we do not have too much zeros) are replaced by half of the minimum positive value
- Result may change if the initial value changes(of course, we want the global minimum of the optimization problem). 




