---
title: "hierarchical_model_v2"
author: "Yi Li"
date: "08/18/2015"
output: pdf_document
---


# 1. Review

Three important elements at location $x$, time $t$.:

- Larval production ($L_{x, t}$), 
- recruitment ($R_{x, t}$),
- adult abundance ($A_{x, t}$),

together with the environment features $E_{x,t}$

Three stages:

- Larval production: 
\[
L_{x, t} = f \left( A_{x, t}, E_{x, t} \right) = p_{x, t} \cdot A_{x, t} \cdot E_{x, t}
\]
- Recruitment
\[
R_{x, t} = k + \log\left(\sum_{i=1}^N L_{i, t} \cdot e^{-\frac{\left(d_{x,i} - \mu_d\right)^2}{\sigma^2_d}} \right) + \beta \cdot E_{x, t} + \epsilon
\]
- Adult abundance
\[
A_{x, t+1} = b \cdot R_{x, t} \cdot E_{x, t+1}
\]

# 2. Update to second order of the environment feature
## 2.1 Larval production: 
\[
\log L_{x, t} = (\log A_{x, t}) (\beta_0+\beta_1E_{x, t}+\beta_2E_{x, t}^2) 
\]

This is equivalent to 
\[
L_{x, t} =  A_{x, t}e^{\beta_0+\beta_1E_{x, t}+\beta_2E_{x, t}^2}
\]

## 2.2 Recruitment
\[
\log R_{x, t} = k + \log\left(\sum_{i=1}^N L_{i, t} \cdot e^{-\frac{\left(d_{x,i} - \mu_d\right)^2}{\sigma^2_d}} \right) + \beta_3E_{x, t}+ \beta_4E_{x, t}^2 + \epsilon
\]

We don't have data for $L_{x,t}$, but we can plug $L_{x,t}$(from 2.1) in the above equation:

\[
\log R_{x, t} = k + \log\left(\sum_{i=1}^N A_{i, t} \cdot e^{\beta_0+\beta_1E_{i, t}+\beta_2E_{i, t}^2-\frac{\left(d_{x,i} - \mu_d\right)^2}{\sigma^2_d}} \right) + \beta_3E_{x, t}+ \beta_4E_{x, t}^2 + \epsilon
\]

## 2.3 Adult abundance
\[
A_{x, t+1} = R_{x, t}e^{\beta_6+\beta_7E_{x, t+1}+\beta_8E_{x, t+1}^2}+mA_{x,t}
\]

# 3. Two stages model
## 3.1 Model 1
\[
\log R_{x, t} = \log\left(\sum_{i=1}^N A_{i, t} \cdot e^{p_4 + p_5E_{i, t} - p_6E_{i, t}^2 + p_7d_{x,i} - p_8d_{x,i}^2} \right) + p_1+ p_2E_{x, t} - p_3E_{x, t}^2 + \epsilon
\]


Note that:

- Unimodality of the environment: coefficient of the second order term is negative
- $argmax(-ax^2+bx)=\frac{b}{2a}$ implies $E^*$ occurs at $\frac{p_5}{2p_6}$ and $\frac{p_2}{2p_3}$
- Dispersal parameters: $\mu_d=\frac{p_7}{2p_8}$, $\sigma_d=\frac{1}{\sqrt{p_8}}$

## 3.2 Adult abundance
\[
A_{x, t+1} = R_{x, t}e^{p_2+p_3E_{x, t+1}+p_4E_{x, t+1}^2}+p_1A_{x,t}
\]

Note that:

- This is the growth from $t$ to $t+1$. may try environment variable $E_{x,t}$ or $E_{x,t+1}$


# 4. Code and Results

## 4.1 load the data

```{r}
X=read.csv("~/Dropbox/model/adults.recruits.merged.csv")
feature_name=c("year","lat","lon","sitenum","cover",
               "filter_chla_mean","filter_sst_mean","filter_upw_mean","mytilus.rec")
X=X[feature_name]
```

## 4.2 derive new variables

- next year abundance: $y=A_{x,t+1}$, target variable


```{r}
nsample=dim(X)[1]
X["ycover"]=NA
for(i in 1:nsample){
  idx=(X[,"sitenum"]==X[i,"sitenum"])&(X[,"year"]==X[i,"year"]+1)
  if(sum(idx)){ # if no next value, will be NA
    X[i,"ycover"]=X[idx,"cover"]
  }
}
```

- next year environment: $E_{x,t+1}$, will be used in model 2

```{r}
X["new_upw"]=NA
X["new_chla"]=NA
X["new_sst"]=NA
for(i in 1:nsample){
  idx=(X[,"sitenum"]==X[i,"sitenum"])&(X[,"year"]==X[i,"year"]+1)  
  if(sum(idx)){ # if no next value, will be NA
    X[i,"new_upw"]=X[idx,"filter_upw_mean"]
    X[i,"new_chla"]=X[idx,"filter_chla_mean"]
    X[i,"new_sst"]=X[idx,"filter_sst_mean"]
    
  }
}
```

Find the complete observations and relabel the site number. Note that we only have one observation for site number 8. And after relabeling, site labels are increaseing with latitude. 

```{r}
X1=X[complete.cases(X),]

lats=subset(X1,year==2001,select = c(lat))[,1]
lons=subset(X1,year==2001,select = c(lon))[,1]
lats=floor(lats*1000)
lons=floor(lons*1000)
# unique identifier of each site, sorted by lats
# using both lat and lon, because there are two sites have the same lat(but different lon)
latlon=paste0(lats,lons) 

for(i in 1:dim(X1)[1]){
  X1[i,"sitenum"]=which(latlon==paste0(floor(X1[i,"lat"]*1000),floor(X1[i,"lon"]*1000)))
}
```

- Target variable in model 1: Recruitment $R_{x,t}$

```{r}
R=X1[,"mytilus.rec"]
R[R==0]=min(R[R!=0])/2
R=log(R)
```

- Target variable in model 2: next year abundance $y=A_{x,t+1}$

```{r}
y=X1[,"ycover"]
y[y==0]=min(y[y!=0])/2 # replace 0 to the minimum positive number
y=log(y) # take log of the species number
```

- Distance Matrix

```{r}
D=dist(X1[X1[,"year"]==2001,"lat"],diag = T,upper = T)
D=as.matrix(D)
for(i in 1:48){
  for(j in 1:48){
    if(i>j){
      D[i,j]=-D[i,j]
    }
  }
}
```

Rescale the environment variables, record the mean and standard deviation. 

```{r}
mean_upw=mean(X1[,"filter_upw_mean"])
sd_upw=sd(X1[,"filter_upw_mean"])
X1[,"filter_upw_mean"]=(X1[,"filter_upw_mean"]-mean_upw)/sd_upw
```

## 4.3 Fit model 1

\[
\log R_{x, t} = \log\left(\sum_{i=1}^N A_{i, t} \cdot e^{p_4 + p_5E_{i, t} - p_6E_{i, t}^2 + p_7d_{x,i} - p_8d_{x,i}^2} \right) + p_1+ p_2E_{x, t} - p_3E_{x, t}^2 + \epsilon
\]

The input of the (maximum) likelihood function, which is equivalent to least square, are

- eight parameters
- data
- target recruitment
- distance matrix

```{r}
m1logl=function(p,X1,R,D){
  n=dim(X1)[1]
  f=rep(0,n)
  for(iobs in 1:n){
    this_ker=0
    this_site=X1[iobs,"sitenum"]
    this_year=X1[iobs,"year"]
    # all the sites that have the same year as the current observation
    Xtemp=subset(X1,year==this_year)
    for(j in dim(Xtemp)[1]){
      sitej=Xtemp[j,"sitenum"]
      this_ker=this_ker+Xtemp[j,"cover"]*
        exp(p[4]+p[5]*Xtemp[j,"filter_upw_mean"]-p[6]*Xtemp[j,"filter_upw_mean"]^2
            +p[7]*D[this_site,sitej]-p[8]*D[this_site,sitej]^2)
    }
    #print(this_ker)
    f[iobs]=log(this_ker)+p[1]+p[2]*X1[iobs,"filter_upw_mean"]-p[3]*X1[iobs,"filter_upw_mean"]^2
  }
  return(sum((R-f)^2))
}
```

```{r, cache=TRUE}
p0=c(0,1,1,0,1,1,1,1)/10
resm1=nlm(m1logl,p0,hessian=T,print.level=0,X1=X1,R=R,D=D,iterlim=1e4,steptol=1e-5)
p1=resm1$estimate
p1
```

```{r}
mu_hat=p1[7]/(2*p1[8])
sigma_hat=1/sqrt(p1[8])
E_in=p1[5]/(2*p1[6])
E_out=p1[2]/(2*p1[3])
mu_hat
sigma_hat
E_in*sd_upw+mean_upw
E_out*sd_upw+mean_upw
```

```{r}
m1pred=function(p,X1,R,D){
  n=dim(X1)[1]
  f=rep(0,n)
  for(iobs in 1:n){
    this_ker=0
    this_site=X1[iobs,"sitenum"]
    this_year=X1[iobs,"year"]
    Xtemp=subset(X1,year==this_year)# all the sites that have the same year as the current observation
    for(j in dim(Xtemp)[1]){
      sitej=Xtemp[j,"sitenum"]
      this_ker=this_ker+Xtemp[j,"cover"]*
        exp(p[4]+p[5]*Xtemp[j,"filter_upw_mean"]-p[6]*Xtemp[j,"filter_upw_mean"]^2
            +p[7]*D[this_site,sitej]-p[8]*D[this_site,sitej]^2)
    }
    #print(this_ker)
    f[iobs]=log(this_ker)+p[1]+p[2]*X1[iobs,"filter_upw_mean"]-p[3]*X1[iobs,"filter_upw_mean"]^2
  }
  return(f)
}
```

Plot the predicted value and the true value, for model 1(recruitment).

```{r}
plot(m1pred(p1,X1,R,D),R)
abline(a=0,b=1,col=2)
```

Relationship between the environment variable and the recruitment, seem decreasing.Thus, I tried several starting points, $t_2$ is always negative. 

```{r}
plot(X1[,"filter_upw_mean"],X1[,"mytilus.rec"])
```


## 4.4 Fit model 2

\[
A_{x, t+1} = R_{x, t}e^{p_2+p_3E_{x, t+1}+p_4E_{x, t+1}^2}+p_1A_{x,t}
\]

```{r}
m2logl=function(p,X1,y){
  n=dim(X1)[1]
  f=rep(0,n)
  for(iobs in 1:n){
    f[iobs]=X1[iobs,"mytilus.rec"]*exp(p[2]+p[3]*X1[iobs,"new_upw"]-p[4]*X1[iobs,"new_upw"]^2)+p[1]*X1[iobs,"cover"]
  }
  return(sum((exp(y)-f)^2))
}
```

```{r, cache=TRUE}
p0=c(1,0,1,1)
resm2=nlm(m2logl,p0,hessian=T,print.level=0,X1=X1,y=y,iterlim=1e4,steptol=1e-5)
p2=resm2$estimate
p2
```

```{r}
E_hat=p2[3]/(2*p2[4])
E_hat*sd_upw+mean_upw
```


```{r}
m2pred=function(p,X1){
  n=dim(X1)[1]
  f=rep(0,n)
  for(iobs in 1:n){
    f[iobs]=X1[iobs,"mytilus.rec"]*exp(p[2]+p[3]*X1[iobs,"new_upw"]-p[4]*X1[iobs,"new_upw"]^2)+p[1]*X1[iobs,"cover"]
  }
  return(f)
}
```

Plot the predicted value and the true value, for model 2(next year abundance).

```{r}
plot(m2pred(p2,X1),X1[,"ycover"])
abline(a=0,b=1,col=2)
```


# 5. A simplified model

If ignore the environment effect in the first Larval production step, and self-increament term $mA_{x,t}$ in the last Adult abundance step, then the model is equivalent to the previous model with second order environment terms, i.e.


\[
y=K+\log(\sum_{i=1}^{n_s}N_{ti}e^{-\frac{(d_{si}-\mu_d)^2}{\sigma_d^2}})+\beta_1e_{chi}+\beta_2e_{sst}+
\beta_3e_{upw}+\beta_4e_{chi}^2+\beta_5e_{sst}^2+
\beta_6e_{upw}^2+\epsilon,
\]


```{r}
library(synchrony)
data(pisco.data)
y=subset(pisco.data,year>2000,select=c(mussel_abund)) # species number
y[y[,1]==0,]=min(y[y[,1]!=0,])/2 # replace 0 to the minimum positive number
y=log(y) # take log of the species number

N=subset(pisco.data,year<2003,select=c(mussel_abund)) # (past) species number
D=dist(pisco.data[1:48,1],diag = T,upper = T)
D=as.matrix(D)
for(i in 1:48){
    for(j in 1:48){
        if(i>j){
            D[i,j]=-D[i,j]
        }
    }
}
E=subset(pisco.data,year<2003,select=c(chl,sst,upwelling)) # Environment variable
D=as.matrix(D)
```

```{r}
logl2=function(t,y,N,D,E){
  n=dim(y)[1]
  f=rep(0,n) # value of the regression function
  yr=rep(0:2,each=48) # 3 years(0-2) used: 00-03 for X, 01-04 for y
  for(j in 1:n){
    Ker=exp(-(D[j%%48+((j%%48)==0)*48,]-t[2])^2/(t[3]^2)) # dispersal kernal (for the 48 sites)
    Nj=N[(yr[j]*48+1):((yr[j]+1)*48),] # number of species for 48 sites
    f[j]=t[1]+log(sum(Ker*Nj))+
      t[4]*E[j,1]+t[5]*E[j,2]+t[6]*E[j,3]+
      t[7]*E[j,1]^2+t[8]*E[j,2]^2+t[9]*E[j,3]^2 # dispersal + three environment terms
  }
  
  return(sum((y[,1]-f)^2)) # return the objective function
}
```

```{r,cache=TRUE}
t0=c(1,0,1,10,10,20,10,10,20)
res2=nlm(logl2,t0,hessian=T,print.level=1,y=y,N=N,D=D,E=E,iterlim=1e4,steptol=1e-5)
t1=res2$estimate
t1
```

```{r}
t1[2]
t1[3]
-t1[6]/(2*t1[9])
```

For the result parameter $t$, note that $\mu_d=t_2$, $\sigma_d=t_3$, $E^*=-\frac{t_6}{2t_9}$.
